---
title: "(More on) What is going on with crime (types) in Phoenix?"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: 
  html_document:
    theme: yeti
    toc: false
    toc_float: false
    collapsed: false
    smooth_scroll: false
---

```{r setup, include=FALSE}

knitr::opts_chunk$set(echo = TRUE, 
                      message = FALSE, 
                      warning = FALSE, 
                      fig.width = 12, 
                      fig.height = 10)

rm(list = ls())

```


```{r, echo = FALSE, eval = TRUE}

# ~~~~~~~~~~~~~~~~~~~~~~~~~~~~~ #
# I run this chunk so that I can go through it and have the tabs show the code.

library( dplyr )
library( tidyr )
library( ggplot2 )
library( forecast )


# Get the data.
url <- "https://www.phoenixopendata.com/dataset/cc08aace-9ca9-467f-b6c1-f0879ab1a358/resource/0ce3411a-2fc6-4302-a33f-167f68608a20/download/crimestat.csv"
crime.data <- read.csv( url, as.is = TRUE, header = TRUE )
crime.data <- na.omit( crime.data )

# Help with formating the text.
library(scales)
n.cases <- dim( crime.data )[1]

# Clean up the dates.
date.vec <- strptime( crime.data$OCCURRED.ON, format="%m/%d/%Y %H:%M" )
crime.data$year   <- format( date.vec, format="%Y" )
crime.data$month  <- format( date.vec, format="%B" )


# Crimes by month.
crimes.by.month <-
  crime.data %>% 
  select( year, month ) %>%   
  filter( year != 2015 ) %>%  
  filter( !is.na( year ) ) %>% 
  group_by( year, month ) %>% 
  summarize( counts = n() ) %>% 
  arrange( match( month, month.name ) ) 
crimes.by.month$month <-factor( crimes.by.month$month,levels = month.name )
crimes.by.month$month <-factor( month.abb[crimes.by.month$month],levels = month.abb )


# Create crimes by type by month object.
crimes.type.by.month <-
  crime.data %>% 
  select( year, month, UCR.CRIME.CATEGORY ) %>%   
  filter( year != 2015 ) %>%  
  filter( !is.na( year ) ) %>% 
  group_by( year, month, UCR.CRIME.CATEGORY ) %>% 
  summarize( counts = n() ) %>% 
  arrange( match( month, month.name ) ) %>% 
  select( !month )
crimes.type.by.month$month <-factor( crimes.type.by.month$month,levels = month.name )
crimes.type.by.month$month <-factor( month.abb[crimes.type.by.month$month],levels = month.abb )


```


<br>

In a prior post, I talked about what is going on with [crime in Phoenix in 2020](https://jacobtnyoung.github.io/R2PhX/projects/PHXcrime_trends.html), noting the considerable decline in crime relative to other years. 

Here, I want to further unpack the trends by focusing on crime types.


<br><br>


### Monthly Counts over the Years {.tabset}

<br>

#### 

<br>

Let's pull the most recent data for **crime incidents** from the site.

The data are reported as UCR crime classifications and have geographic information (block address, zip) as well as the date and time of the incident. As of `r format(Sys.time(), '%B, %Y')`, there were `r comma(n.cases)` crime incidents with complete data from 11/2015 to within a week of the current date.   

Let's start by looking at the monthly count of incidents over all crime types.
  
<br>

```{r, echo=FALSE, eval=TRUE}

crimes.by.month %>% 
  ggplot( aes( month, counts, group = 1 ) ) +
  geom_point( ) +
  geom_line( ) +
  facet_grid( ~ year , scales="free" )

```

HERE


```{r, echo=FALSE, eval=TRUE}

crimes.type.by.month %>% 
  ggplot( aes( month, counts, group = 1 ) ) +
  geom_point( ) +
  geom_line( ) +
  facet_grid( UCR.CRIME.CATEGORY ~ year, scales="free" )

```


###

<br>

<p align="center">
[Back to R 2 Phoenix page](https://jacobtnyoung.github.io/R2PhX/)
</p>

<br>

###### ***Last updated `r format(Sys.time(), '%d %B, %Y')`***