---
title: "Are officer involved shootings increasing in Phoenix? (and other questions...)"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: 
  html_document:
    theme: yeti
    toc: false
    toc_float: false
    collapsed: false
    smooth_scroll: false
---

```{r setup, include=FALSE}

knitr::opts_chunk$set( echo = TRUE, 
                      message = FALSE, 
                      warning = FALSE, 
                      fig.width = 12, 
                      fig.height = 10 )

rm( list = ls() )

```

<br>

```{r, echo = FALSE, eval = TRUE}

url <- "https://www.phoenixopendata.com/dataset/1e472b1f-80c1-4686-b80f-e930616f8580/resource/8f8c1739-b905-407f-9eb6-6b10db55b931/download/ois.csv"
shooting.data <- read.csv( url, as.is = TRUE, header = TRUE )
shooting.data <- na.omit( shooting.data )

library( scales )
n.cases <- dim( shooting.data )[1]

library( dplyr )
library( tidyr )
library( ggplot2 )
library( forecast )


```  

The behavior of police officers has been a major event in 2020. So, we might wonder whether officer involved shootings have become more common (or rare?) in Phoenix over the last few years. We can try and answer this question by examining the [officer involved shootings](https://www.phoenixopendata.com/dataset/officer-involved-shooting-incidents) data from the [City of Phoenix](https://www.phoenix.gov/) open [data](https://www.phoenixopendata.com/) portal. Let's get to it!

<br>

###  {.tabset}

#### About the Data {.tabset}

The dataset is a CSV file containing all [officer involved shooting incidents](https://www.phoenixopendata.com/dataset/officer-involved-shooting-incidents) by Phoenix Police Department officers from January 1, 2017 to a week prior to the current date. Accidental discharges, discharges at animals, and discharges at objects where there was not an active threat by a subject are not included in this data set.

As of `r format(Sys.time(), '%B, %Y')`, there were `r comma( n.cases )` crime incidents with complete data from January 2017 to within a week of the current date.  

<br>

The dataset contains information about:  

  + *When* the incident occurred (e.g. `DATE`, `TIME`, `HOUR`)  

  + *Where* the incident occurred (e.g. `LOCATION`, `PRECINCT`, `COUNCIL_DIST`)  
  
  + The *suspect*, denoted by the `SP_` prefix (e.g. `SP_RACE`, `SP_GENDER`)  
  
  + The *officer(s)* involved, deonted by the `PO_` prefix (e.g. `PO_RACE`, `PO_ETHNICITY`)  
  
  + And the *outcome*, such as the injury level (i.e. `SP_INJURY_LEVEL`) and whether there was a fatality (i.e. `FATALITY_FLG`)  
  
<br>


#### Getting the data (code)  

```{r, echo = TRUE, eval = FALSE}

# set the url where the data are located.
url <- "https://www.phoenixopendata.com/dataset/1e472b1f-80c1-4686-b80f-e930616f8580/resource/8f8c1739-b905-407f-9eb6-6b10db55b931/download/ois.csv"

# pull in the csv file.
shooting.data <- read.csv( url, as.is = TRUE, header = TRUE )

# drop cases missing on date.
shooting.data <- na.omit( shooting.data )

# take a look at the data.
head( shooting.data )  

``` 


#### Preprocessing the data (code)  

Now that the data are in the workspace, let's clean up the dates using the `strptime()` and `format()` functions here.

```{r, echo = TRUE, eval = TRUE}

# The date and time variable is a character string.
head( shooting.data$DATE )
is.character( shooting.data$DATE )

# Since the month variable is in an unusual format, we need to convert the date string to a date format code. Then pull off the month.
date.vec <- strptime( shooting.data$DATE, format="%d-%b-%y" )

# Now, let's use the format() function to create the month.
shooting.data$month  <- format( date.vec, format="%B" )

```


<br>

###  {.tabset}

#### Monthly Incidents

<br>

As a first step, let's take a look at the monthly occurrence of officer involved shootings. The plot shows the monthly count of officer involved shootings from January, 2017 to `r format(Sys.time(), '%B, %Y')`. 

<br>

```{r, echo = FALSE, eval = TRUE}

# Create an object that is monthly counts and sorted by year. 
shootings.by.month <- 
  shooting.data %>% 
  select( YEAR, month ) %>%   
  group_by( YEAR, month ) %>% 
  summarize( counts = n() ) %>% 
  spread( YEAR, counts ) %>% 
  arrange( match( month, month.name ) ) %>% 
  select( !month )

# Replace missing values with zero.
shootings.by.month[ is.na( shootings.by.month ) == TRUE ] <- 0

# Use the ts() function to create a time series object.
monthly.shootings.by.year <- ts(
  matrix( as.matrix( shootings.by.month ), ncol = 1 ), start=c( 2017, 1 ), end=c( 2020, 12 ), frequency=12
)

# Finally, let's take a look using ggseasonplot().
library( ggplot2 )
library( forecast )

# Now, plot it.
monthly.shootings.by.year %>% 
  ggseasonplot(
    year.labels=FALSE,
    main = "Plot of Monthly Officer Involved Shootings by Years for Phoenix" ) + 
  scale_y_continuous(label = comma) +
  theme_gray() 

```

<br>

The plot shows several interesting patterns:  

  + First, there are months throughout the years in which there are no shootings. In fact, there are `r length( which( shootings.by.month == 0 ) )` months in the data in which there are no shootings.  
  
  + Second, April 2018 sticks out as there were `r max( shootings.by.month )` in that month. 
  
  + Third, 2020 does not deviate from the other years in the data. Specifically, there were `r sum( shootings.by.month$"2020" )` officer involved shootings in 2020. For 2017, 2018, and 2019, there were `r sum( shootings.by.month$"2017" )`, `r sum( shootings.by.month$"2019" )`, and `r sum( shootings.by.month$"2019" )` officer involved shootings, respectively. 


#### (code)

<br>

```{r, echo = TRUE, eval = FALSE}

# Create an object that is monthly counts and sorted by year. 
shootings.by.month <- 
  shooting.data %>% 
  select( YEAR, month ) %>%   
  group_by( YEAR, month ) %>% 
  summarize( counts = n() ) %>% 
  spread( YEAR, counts ) %>% 
  arrange( match( month, month.name ) ) %>% 
  select( !month )

# Replace missing values with zero.
shootings.by.month[ is.na( shootings.by.month ) == TRUE ] <- 0

# Use the ts() function to create a time series object.
monthly.shootings.by.year <- ts(
  matrix( as.matrix( shootings.by.month ), ncol = 1 ), start=c( 2017, 1 ), end=c( 2020, 12 ), frequency=12
)

# Finally, let's take a look using ggseasonplot().
library( ggplot2 )
library( forecast )

# Now, plot it.
monthly.shootings.by.year %>% 
  ggseasonplot(
    year.labels=FALSE,
    main = "Plot of Monthly Officer Involved Shootings by Years for Phoenix" ) + 
  scale_y_continuous(label = comma) +
  theme_gray() 

```

<br>

###  {.tabset}

#### Suspect Characteristics

<br>

Additionally, we can examine the characteristics of the individuals who are shot. In the data, there are several variables for "suspect characteristics".  

There are a few interesting features of the data:  

  + Of the `r sum( shootings.by.month)` shootings, `r table( shooting.data$FATALITY_FLG)[1]` were fatal, a proportion of `r table( shooting.data$FATALITY_FLG)[1] / sum( shootings.by.month)`.
  
  + Whites account for the majority of cases, with a proportion of `r round( ( table( shooting.data$SP_RACE[shooting.data$SP_RACE == "White"] ) / sum( shootings.by.month ) ), 2 )`. 

The plot below shows the frequency distribution (i.e. counts) of shootings by age, outcome (where red is a fatality), and faceted by whether the suspect is White or Non-White. 

<br>


```{r, echo = FALSE, eval = TRUE}

# Create a plot with race, age, and the outcome.
ggplot(
  shooting.data %>% 
  select( SP_RACE, SP_AGE_AT_INC, FATALITY_FLG ) %>% 
  mutate( sp_wnw = ifelse( SP_RACE != "White", "Non-White", "White" ) ) %>%
  mutate( SP_AGE_AT_INC = replace( SP_AGE_AT_INC, SP_AGE_AT_INC == 120, 12 ) ) %>% 
  group_by( sp_wnw, SP_AGE_AT_INC, FATALITY_FLG ) %>% 
  summarize( counts = n() ) ) +
 geom_col( aes( SP_AGE_AT_INC, counts, fill = FATALITY_FLG ) ) +
 facet_wrap( ~sp_wnw ) +
 labs( x = "Suspect Age" ) +
 labs( fill = "Outcome of Shooting" ) +
 scale_fill_manual(values = c("#ca0222", "#02677b")) +
 labs( title = "Plot of Suspect Age, Suspect Race, and Outcome for Officer Involved Shootings in Phoenix" ) +
 theme_gray() 

```

<br>


<br>

#### (code)

<br>

```{r, echo = TRUE, eval = FALSE}

# Create a plot with race, age, and the outcome.
ggplot(
  shooting.data %>% 
  select( SP_RACE, SP_AGE_AT_INC, FATALITY_FLG ) %>% 
  mutate( sp_wnw = ifelse( SP_RACE != "White", "Non-White", "White" ) ) %>%
  mutate( SP_AGE_AT_INC = replace( SP_AGE_AT_INC, SP_AGE_AT_INC == 120, 12 ) ) %>% 
  group_by( sp_wnw, SP_AGE_AT_INC, FATALITY_FLG ) %>% 
  summarize( counts = n() ) ) +
 geom_col( aes( SP_AGE_AT_INC, counts, fill = FATALITY_FLG ) ) +
 facet_wrap( ~sp_wnw ) +
 labs( x = "Suspect Age" ) +
 labs( fill = "Outcome of Shooting" ) +
 scale_fill_manual(values = c("#ca0222", "#02677b")) +
 labs( title = "Plot of Suspect Age, Suspect Race, and Outcome for Officer Involved Shootings in Phoenix" ) +
 theme_gray() 

```

###

<br>

<p align="center">
[Back to R 2 Phoenix page](https://jacobtnyoung.github.io/R2PhX/)
</p>

<br>

###### ***Last updated `r format(Sys.time(), '%d %B, %Y')`***
